The Assembly of Black Hole Mass and Luminosity Functions of High-redshift Quasars\\via Multiple Accretion Episodes

### 2022-12-28 ms.tex
review1228; publicated ref updated

\vspace{5mm}
\section{Introduction} \label{sec:intro}
The growth of massive BHs in active galactic nuclei (AGN) is elegantly related to their luminous accretion phases
across the cosmic time \citep{1982MNRAS.200..115S}.
The cosmic BH accretion history and their radiative efficiency are well constrained by 
comparing the mass density of supermassive black holes (SMBHs) in the local universe and the mass accreted onto BHs inferred from
the integration of the QLF based on multi-wavelength observations
\citep[e.g.][]{1971ApJ...170..223C,1992MNRAS.259..725S,2002MNRAS.335..965Y,2004MNRAS.351..169M,2008MNRAS.388.1011M,2004MNRAS.354.1020S,
2009ApJ...690...20S,2014MNRAS.439.2736D,2014ApJ...786..104U,2017A&A...600A..64T,2022arXiv220607357S}.

@@

The shapes of the QLF and BHMF contain key information on the physics characterizing the radiative 
and accretion processes of high-$z$ SMBHs, as well as their seeding mechanisms in principle.
Two approaches have been utilized to model those distribution functions.  
The first is semi-analytical modeling, in which BH seed formation, gas accretion, and BH mergers associated 
with the hierarchical growth of the parent dark matter (DM) halos are taken into account in a simplified way. 
This is an effective way of examining the statistical properties of the early BH population and making predictions 
that can be directly compared with the observed QLF and BHMF
\citep[e.g.,][]{1998ApJ...503..505H,2010ApJ...718..231S,2018MNRAS.474.1995R,2018MNRAS.481.3278R,
2021MNRAS.508.2706Y,2021ApJ...910L..11K,2022MNRAS.511..616T,2022arXiv220714689O}. 
However, due to the limitation of capturing the detailed properties of the relevant physics, 
a large number of model parameters need to be calibrated based on the observed distribution functions 
of lower-$z$ quasar populations \citep[e.g.,][]{2007ApJ...669...45H}
and empirical correlations between the SMBH mass and their host galaxy properties seen
in the local universe \citep{2013ARA&A..51..511K}.
The second approach is more phenomenological, using extrapolation of the low-$z$ QLF evolution with 
fitting function forms \citep[e.g.,][]{2019MNRAS.488.1035K,2020MNRAS.495.3252S,2022arXiv220702233F}.
This method also allows us to forecast the higher-$z$ QLF and bring insights for future surveys.
Nevertheless, those semi-analytical and phenomenological approaches introduce numerous physical parameters
and make the essential mechanisms reproducing the observed results elusive.
Therefore, it is required to construct a theoretical model characterized by a minimum number of parameters
but capable of dealing with the essential physical processes.

To determine the initial conditions of the early BH assembly,
the formation pathway of seed BHs has been extensively investigated by numerical simulations and semi-analytical studies
\citep{2006MNRAS.370..289B,2009MNRAS.396..343R,2009ApJ...696.1798T,2012MNRAS.422.2051N,2014ApJ...781...60H,
2015MNRAS.448..568H,2018MNRAS.474.3825V,2021MNRAS.506..613S,2022arXiv220614459T,2022arXiv220505717B}.
Massive star formation episodes in the early universe is substantially modulated by external environmental effects such as
(i) H$_2$ photo-dissociating irradiation from nearby galaxies
\citep{2001ApJ...546..635O,2002ApJ...569..558O,2003Natur.425..812B,2010MNRAS.402.1249S,2014MNRAS.445..544S,2014MNRAS.445..107V,2016ApJ...832..134C},
(ii) supersonic baryonic streaming motion that delays gas collapse in halos
\citep{2012MNRAS.424.1335F, 2014MNRAS.439.1092T, 2018ApJ...855...17H,2017MNRAS.471.4878S,2018MNRAS.479.4017I},
and (iii) dynamical heating caused by frequent halo mergers and turbulence injection led by cold accretion flows
\citep{2003ApJ...592..645Y,2010Natur.466.1082M,2015ApJ...810...51M,2019Natur.566...85W,2022Natur.607...48L}.
Those effects keep isothermal collapse ($T\sim 8000~{\rm K}$) of massive gas at high accretion rates of $\gtrsim 0.1~\Msunyr$
without vigorous fragmentation \citep{2014MNRAS.445L.109I,2015MNRAS.446.2380B,2016PASA...33...51L},
and allow supermassive stars with $\sim 10^5~\Msun$ (presumably seed BHs) to form at the halo centers 
\citep{2013ApJ...778..178H,2013A&A...558A..59S,2019PASA...36...27W,2022arXiv220614459T}.
Although the conditions required for heavy seed formation are thought to be too stringent to be realized in the typical regions of the early universe
\citep{2008MNRAS.391.1961D,2009ApJ...695.1430A,2015MNRAS.450.4350I},
recent studies by \citet{2021MNRAS.503.5046L} and \citet{2021ApJ...917...60L} found that the formation of heavy seeds
accelerates in rare, overdense regions of the universe, where non-linear galaxy clustering boosts the irradiation intensity
and frequency of halo mergers.
\blue{In such massive halos, intense cold streams through cosmic webs and supersonic turbulent motion promote the formation
of massive gas clouds that collapse into supermassive stars and seed BHs \citep{2009Natur.457..451D,2012MNRAS.422.2539I,2022Natur.607...48L}.}
The mass distribution of seed BHs is expected to be substantially top-heavy, extending the upper mass to $\gtrsim 10^{4-5}~\Msun$
\citep{2021ApJ...917...60L, 2022arXiv220614459T}.
\blue{
Metal enrichment of the progenitor halos due to internal and external star formation activity would regulate the formation of 
seed BHs and alter the characteristic of the mass distribution function (see also a quantitative argument on the low efficiency of enrichment in 
\citealt{2021ApJ...917...60L}).
Further studies with hydrodynamic simulations are required to improve our understanding of the efficiency of metal enrichment in quasar progenitor halos
\citep[e.g.,][]{2018MNRAS.475.4378C}.
}
%As a result, \citet{2021ApJ...917...60L} predicted a top-heavy mass distribution of seed BHs ranging from $10^2-10^5~\Msun$.

@@

In addition, we consider baryonic streaming motion relative to DM produced in the epoch of cosmic recombination at $z_{\rm rec} \simeq 1100$.
This effect also delays gas collapse and star formation in protogalaxies through injection of kinetic energy into gas
\citep[e.g.,][]{2012MNRAS.424.1335F, 2014MNRAS.439.1092T, 2017Sci...357.1375H,2019MNRAS.484.3510S,2021ApJ...917...60L}.
The amplitude of the streaming velocity is set to $\vbsm = 0$ and $\vbsm=1~\sigma_{\rm bsm}$, where $\sigma_{\rm bsm}=30~{\rm km~s}^{-1} (1+z)/(1+z_{\rm rec})$
is the root-mean-square speed at a redshift of $z$ \citep{2010PhRvD..82h3520T}.
We calculate the effective sound speed of gas as $c_{\rm eff} = \{c_{\rm s}^2+v_{\rm tur}^2/3+\left(\alpha_0 \vbsm \right)^2 \}^{1/2}$ at the center of each progenitor halo,
where $c_{\rm s}$ is the thermal sound speed, $v_{\rm tur}$ is the turbulent velocity of gas (i.e., the specific kinetic energy accumulated through halo mergers), and
the coefficient is set to $\alpha_0=1$ \footnote{The value of $\alpha_0=1.0$ is motivated by cosmological simulations of massive primordial stars
under fast streaming velocities \citep{2017Sci...357.1375H,2019MNRAS.484.3510S}.
Note that our previous study in \citet{2021ApJ...917...60L} adopted $\alpha_0=4.7$ \citep{2018ApJ...855...17H}.}.
When the gas core becomes gravitationally unstable, the evolution of the gas density profile is well described by 
the Penston-Larson self-similar solution \citep{1969MNRAS.144..425P,1969MNRAS.145..271L}. 
The mass accretion rate from the envelope can be written as $\dot{M}=c_{\rm eff}^3/G$,
where $G$ is the gravitational constant.

\blue{
At the vicinity of an accreting protostar, an accretion disk forms due to angular momentum of the inflowing material from large scales
\citep{2010MNRAS.403...45S,2011Sci...331.1040C,2011ApJ...737...75G}.}
The episodic nature of disk accretion decreases the time-averaged accretion rate onto the protostar as $\Mdot_{\star}=\eta \Mdot$, 
where we adopt $\eta=0.3$ as the conversion efficiency \blue{owing to angular momentum of the accreting flow} \citep{2016MNRAS.459.1137S, 2022arXiv220614459T}. 
%\blue{
%  (Toyouchi+ find 0.4 actually.)
%}
The size evolution of an accreting protostar and the radiative feedback strength depend sensitively on whether
the accretion rate becomes higher than a critical value of $\dot{M}_{\mathrm{crit}}=0.04~\Msunyr$ 
\citep{2001ApJ...561L..55O,2013ApJ...778..178H,2013A&A...558A..59S,2015MNRAS.452..755S,2018MNRAS.474.2757H}.
For $\dot{M}_\star > \Mdot_\mathrm{crit}$, the stellar envelope inflates owing to rapid entropy injection through accreting matter,
keeping its surface temperature as low as $T_\mathrm{eff} \simeq 5000$ K. 
Since stellar UV radiation is hardly emitted from the cold surface, the accretion flow efficiently feeds the central star without being 
impeded by radiative feedback.
As a result, the stellar mass reaches $\sim 10^5~\Msun$ but is limited by the general-relativistic instability to
%
\begin{equation}
 M_{\star, \mathrm{GR}} \simeq\left[0.83~\log \left(\frac{\dot{M}_{\star}}{\Msunyr}\right)+2.48\right] \times 10^{5} ~\Msun,
\end{equation}
%
in the range of $\dot{M}_\star\gtrsim 0.1~\Msunyr$ \citep{2016PhRvD..94b1501S,2017ApJ...842L...6W,2019PASA...36...27W}.
On the other hand, for $\dot{M}_\star \leq \Mdot_\mathrm{crit}$, the protostar begins to contract by loosing its thermal energy via radiative diffusion
and increases its surface temperature to $T_\mathrm{eff} \sim 10^5$ K.
Hence, intense stellar UV radiation heats the disk surface and launches mass outflows, preventing mass supply from the disk to the star
\citep{2008ApJ...681..771M,2011Sci...334.1250H}. 
In this case, the stellar mass is determined by the balance between mass accretion and mass loss via photoevaporation \citep{2013ApJ...773..155T}.
The feedback-regulated mass can be approximated as
\footnote{A recent study by \citet{2022arXiv220614459T} provides a more sophisticated prescription of the feedback-regulated stellar mass based on their radiation hydrodynamical simulations.}
%
\begin{equation}
M_{\star, \mathrm{fb}} \simeq 2.9 \times 10^{3} \Msun\left(\frac{\dot{M}_{\star}}{0.01 M_{\odot} \mathrm{yr}^{-1}}\right),
\end{equation}
%
see more details in \cite{2021ApJ...917...60L}.
For an intermediate accretion rate between $\Mdot_\mathrm{crit}$ and $0.1~\Msunyr$, 
we perform logarithmic interpolation in the $\Mdot_\star - M_\star$ plane to smoothly connect 
$M_{\star, \mathrm{fb}}$ and $M_{\star, \mathrm{GR}}$ at the boundaries.
At the end of the stellar lifetime, those massive primordial stars result in BHs without significant mass loss 
via stellar winds \citep{2003ApJ...591..288H,2015MNRAS.451.4086S}.
\blue{One important caveat is the effect of stellar rotation on the structure evolution.
Namely, the mass loss rate of a rotating massive star via winds would be enhanced by the centrifugal force on the surface
or suppressed by efficient mixing of the interior structure by meridional circulation \citep[see][references therein]{2008A&A...489..685E,2012A&A...542A.113Y}.
Although those effects in the high mass regime are poorly understood yet, full general-relativistic simulations of the gravitational collapse of a rotating supermassive star 
show that most of the stellar mass is eventually swallowed by the newly born BH, ejecting only $\sim 10\%$ of the mass \citep{2002ApJ...572L..39S,2016PhRvD..94b1501S}.}
Therefore, we assume that the mass of a remnant BH is equal to that of its stellar progenitor.
%\footnote{
%\blue{Stellar rotation is expected to induce various effects on the structure evolution; namely, the mass loss rate via winds would be enhanced by the centrifugal force on the surface
%or suppressed by efficient mixing of the interior structure by meridional circulation \citep[see][references therein]{2008A&A...489..685E,2012A&A...542A.113Y}.
%Although those effects in the high mass regime are poorly understood yet, full general-relativistic simulations of the gravitational collapse of a rotating supermassive star 
%found that most of the stellar mass is eventually swallowed by the newly born BH, ejecting only $\sim 10\%$ of the mass \citep{2002ApJ...572L..39S,2016PhRvD..94b1501S}.}}.

@@

\vspace{2mm}
\subsection{BH mass growth}
\label{sec:model}
With BH seeds planted, we study the evolution of the BH mass distribution until $z=6$. 
The growth model of each BH is characterized by a minimal number of free parameters, 
giving the accretion rate by
%
\begin{equation}
  \label{eq:mdot}
  \Mdot_\bullet = \lambda f(\Mbh) \Mdot_\mathrm{Edd} ,
\end{equation}
where $\lambda$ is the ratio of the quasar bolometric luminosity to its Eddington luminosity $L_{\rm Edd}$,
% $\lambda \equiv L/L_\mathrm{Edd}$ 
and $\Mdot_\mathrm{Edd} \equiv L_{\mathrm{Edd}}/\eta_0 c^2$ is the Eddington accretion rate with a radiative efficiency of $\eta_0=0.1$ \citep{1973A&A....24..337S},
which is consistent with the value obtained from the Soltan-Paczy{\'n}ski argument \citep[e.g.,][]{2002MNRAS.335..965Y,2010ApJ...725..388C}.
Based on studies on the mass dependence of the radiative efficiency for AGNs at $z\lesssim$ 3 
\citep{2008MNRAS.390..561C,2012ApJ...749..187L,2014ApJ...786..104U}, 
we introduce a similar functional form of
%
\begin{equation}
\label{eq:f_M}
f(\Mbh) = \frac{2}{1+\left(\Mbh /M_{\bullet,c} \right)^\delta}, 
\end{equation}
%
where $M_{\bullet,c}=10^8~\Msun$ is adopted
\footnote{
A series of radiation hydrodynamical simulations for BH accretion
suggest that the gas supplying rate from galactic scales into the nuclear scale is not high enough to sustain 
super-Eddington accretion when the BH mass is higher than the characteristic value \citep{2021ApJ...907...74T}.}.
With a positive value of $\delta$, the BH growth at the high mass end is suppressed,
while the growth speed for less massive BHs is accelerated.
In the limit of $\delta \ll 1$, where $f(\Mbh) \simeq 1$ is nearly independent of $M_{\bullet,c}$,
the model in Eq.~(\ref{eq:f_M}) reproduces an exponential growth with an $e$-folding timescale of
$t_{\rm S} =  \Mbh/\Mdot_{\mathrm{Edd}} \approx 45$ Myr (the so-called Salpeter timescale; \citealt{1964ApJ...140..796S}).
\blue{
Note that we do not explicitly consider the energy loss with radiation in Eq.~(\ref{eq:mdot}), neglecting a factor of $1-\eta_0=0.9$.
This small deviation from unity can be absorbed by the uncertainty due to the functional form of $f(M_\bullet)$ and does not bring 
a significant impact on the results discussed below.
}

@@

% obs fraction 
A widely-accepted mechanism causing quasar obscuration is extinction by dusty tori and dense gas clouds 
in the circumnuclear region, hence tightly related to gas fueling and feedback processes from accreting BHs 
\citep[see][for a review]{2018ARA&A..56..625H}.
In addition, dense gas clouds at larger galactic scales may have a significant contribution to quasar obscuration at 
higher redshifts \citep{2020MNRAS.495.2135N}.
%and the obscured fraction is expected to increase with redshift \citep[]{2022arXiv220603508G}. 
%However, we adopt the prescription of the obscured fraction constrained from quasar observations at $z\lesssim 5$ 
However, the physical origin and conditions for high-$z$ quasar obscuration have been poorly understood yet.
Observationally, the obscured fraction of AGNs with hydrogen column density of $N_{\rm H}>10^{22}~{\rm cm}^{-2}$
can be estimated from spectral analyses in the X-ray band
\citep[e.g.,][]{2003ApJ...598..886U,2007A&A...463...79G,2008A&A...490..905H}. 
The obscured fraction is nearly $\gtrsim 80\%$ at $L_{\rm X}<10^{43}~{\rm erg~s}^{-1}$ and 
decreases with the X-ray luminosity \citep{2014ApJ...786..104U,2014MNRAS.437.3550M}.
The exact function shape still remains under debate, but the overall dependence on the X-ray luminosity holds.
Moreover, the obscured fraction is found to increase gradually toward higher redshifts and is saturated at $z\gtrsim 2-3$
\citep{2008A&A...490..905H,2014ApJ...786..104U,2014MNRAS.437.3550M,2018MNRAS.473.2378V,2022arXiv220603508G}.

@@

From another point of view, the discrepancy between the UV and X-ray QLFs at the faint end would be explained by the fact that 
only bright quasars outshining their host galaxies can be identified as point-like sources in the rest-UV quasar selection 
(\citetalias{2018ApJ...869..150M}; \citealt{2020MNRAS.495.2135N,2020MNRAS.494.1771A,2021MNRAS.502.2757O,2021MNRAS.502..662B,2022arXiv220712282K}).
However, at the faint end of $\Muv\gtrsim -24$ mag, extended sources such as the brightest galaxies at $z\sim 6$ 
dominate over quasars in number at the same UV magnitude \citep{2022ApJS..259...20H}.
Thus, the current observations might miss quasars embedded in extended galaxy populations. 
We leave this issue as an important caveat in high-$z$ quasar observations.

>>>>

\vspace{5mm}
\section{Introduction} \label{sec:intro}
The growth of massive BHs in active galactic nuclei (AGN) is elegantly related to their luminous accretion phases
across the cosmic time \citep{1982MNRAS.200..115S}.
The cosmic BH accretion history and their radiative efficiency are well constrained by 
comparing the mass density of supermassive black holes (SMBHs) in the local universe and the mass accreted onto BHs inferred from
the integration of the QLF based on multi-wavelength observations
\citep[e.g.][]{1971ApJ...170..223C,1992MNRAS.259..725S,2002MNRAS.335..965Y,2004MNRAS.351..169M,2008MNRAS.388.1011M,2004MNRAS.354.1020S,
2009ApJ...690...20S,2014MNRAS.439.2736D,2014ApJ...786..104U,2017A&A...600A..64T,2022ApJ...934...66S}.

@@

The shapes of the QLF and BHMF contain key information on the physics characterizing the radiative 
and accretion processes of high-$z$ SMBHs, as well as their seeding mechanisms in principle.
Two approaches have been utilized to model those distribution functions.  
The first is semi-analytical modeling, in which BH seed formation, gas accretion, and BH mergers associated 
with the hierarchical growth of the parent dark matter (DM) halos are taken into account in a simplified way. 
This is an effective way of examining the statistical properties of the early BH population and making predictions 
that can be directly compared with the observed QLF and BHMF
\citep[e.g.,][]{1998ApJ...503..505H,2010ApJ...718..231S,2018MNRAS.474.1995R,2018MNRAS.481.3278R,
2021MNRAS.508.2706Y,2021ApJ...910L..11K,2022MNRAS.511..616T,2022arXiv220714689O}. 
However, due to the limitation of capturing the detailed properties of the relevant physics, 
a large number of model parameters need to be calibrated based on the observed distribution functions 
of lower-$z$ quasar populations \citep[e.g.,][]{2007ApJ...669...45H}
and empirical correlations between the SMBH mass and their host galaxy properties seen
in the local universe \citep{2013ARA&A..51..511K}.
The second approach is more phenomenological, using extrapolation of the low-$z$ QLF evolution with 
fitting function forms \citep[e.g.,][]{2019MNRAS.488.1035K,2020MNRAS.495.3252S,2022ApJ...938...25F}.
This method also allows us to forecast the higher-$z$ QLF and bring insights for future surveys.
Nevertheless, those semi-analytical and phenomenological approaches introduce numerous physical parameters
and make the essential mechanisms reproducing the observed results elusive.
Therefore, it is required to construct a theoretical model characterized by a minimum number of parameters
but capable of dealing with the essential physical processes.

To determine the initial conditions of the early BH assembly,
the formation pathway of seed BHs has been extensively investigated by numerical simulations and semi-analytical studies
\citep{2006MNRAS.370..289B,2009MNRAS.396..343R,2009ApJ...696.1798T,2012MNRAS.422.2051N,2014ApJ...781...60H,
2015MNRAS.448..568H,2018MNRAS.474.3825V,2021MNRAS.506..613S,2023MNRAS.518.1601T,2022MNRAS.516..138B}.
Massive star formation episodes in the early universe is substantially modulated by external environmental effects such as
(i) H$_2$ photo-dissociating irradiation from nearby galaxies
\citep{2001ApJ...546..635O,2002ApJ...569..558O,2003Natur.425..812B,2010MNRAS.402.1249S,2014MNRAS.445..544S,2014MNRAS.445..107V,2016ApJ...832..134C},
(ii) supersonic baryonic streaming motion that delays gas collapse in halos
\citep{2012MNRAS.424.1335F, 2014MNRAS.439.1092T, 2018ApJ...855...17H,2017MNRAS.471.4878S,2018MNRAS.479.4017I},
and (iii) dynamical heating caused by frequent halo mergers and turbulence injection led by cold accretion flows
\citep{2003ApJ...592..645Y,2010Natur.466.1082M,2015ApJ...810...51M,2019Natur.566...85W,2022Natur.607...48L}.
Those effects keep isothermal collapse ($T\sim 8000~{\rm K}$) of massive gas at high accretion rates of $\gtrsim 0.1~\Msunyr$
without vigorous fragmentation \citep{2014MNRAS.445L.109I,2015MNRAS.446.2380B,2016PASA...33...51L},
and allow supermassive stars with $\sim 10^5~\Msun$ (presumably seed BHs) to form at the halo centers 
\citep{2013ApJ...778..178H,2013A&A...558A..59S,2019PASA...36...27W,2023MNRAS.518.1601T}.
Although the conditions required for heavy seed formation are thought to be too stringent to be realized in the typical regions of the early universe
\citep{2008MNRAS.391.1961D,2009ApJ...695.1430A,2015MNRAS.450.4350I},
recent studies by \citet{2021MNRAS.503.5046L} and \citet{2021ApJ...917...60L} found that the formation of heavy seeds
accelerates in rare, overdense regions of the universe, where non-linear galaxy clustering boosts the irradiation intensity
and frequency of halo mergers.
\blue{In such massive halos, intense cold streams through cosmic webs and supersonic turbulent motion promote the formation
of massive gas clouds that collapse into supermassive stars and seed BHs \citep{2009Natur.457..451D,2012MNRAS.422.2539I,2022Natur.607...48L}.}
The mass distribution of seed BHs is expected to be substantially top-heavy, extending the upper mass to $\gtrsim 10^{4-5}~\Msun$
\citep{2021ApJ...917...60L, 2023MNRAS.518.1601T}.
\blue{
Metal enrichment of the progenitor halos due to internal and external star formation activity would regulate the formation of 
seed BHs and alter the characteristic of the mass distribution function (see also a quantitative argument on the low efficiency of enrichment in 
\citealt{2021ApJ...917...60L}).
Further studies with hydrodynamic simulations are required to improve our understanding of the efficiency of metal enrichment in quasar progenitor halos
\citep[e.g.,][]{2018MNRAS.475.4378C}.
}
%As a result, \citet{2021ApJ...917...60L} predicted a top-heavy mass distribution of seed BHs ranging from $10^2-10^5~\Msun$.

@@

In addition, we consider baryonic streaming motion relative to DM produced in the epoch of cosmic recombination at $z_{\rm rec} \simeq 1100$.
This effect also delays gas collapse and star formation in protogalaxies through injection of kinetic energy into gas
\citep[e.g.,][]{2012MNRAS.424.1335F, 2014MNRAS.439.1092T, 2017Sci...357.1375H,2019MNRAS.484.3510S,2021ApJ...917...60L}.
The amplitude of the streaming velocity is set to $\vbsm = 0$ and $\vbsm=1~\sigma_{\rm bsm}$, where $\sigma_{\rm bsm}=30~{\rm km~s}^{-1} (1+z)/(1+z_{\rm rec})$
is the root-mean-square speed at a redshift of $z$ \citep{2010PhRvD..82h3520T}.
We calculate the effective sound speed of gas as $c_{\rm eff} = \{c_{\rm s}^2+v_{\rm tur}^2/3+\left(\alpha_0 \vbsm \right)^2 \}^{1/2}$ at the center of each progenitor halo,
where $c_{\rm s}$ is the thermal sound speed, $v_{\rm tur}$ is the turbulent velocity of gas (i.e., the specific kinetic energy accumulated through halo mergers), and
the coefficient is set to $\alpha_0=1$ \footnote[1]{The value of $\alpha_0=1.0$ is motivated by cosmological simulations of massive primordial stars
under fast streaming velocities \citep{2017Sci...357.1375H,2019MNRAS.484.3510S}.
Note that our previous study in \citet{2021ApJ...917...60L} adopted $\alpha_0=4.7$ \citep{2018ApJ...855...17H}.}.
When the gas core becomes gravitationally unstable, the evolution of the gas density profile is well described by 
the Penston-Larson self-similar solution \citep{1969MNRAS.144..425P,1969MNRAS.145..271L}. 
The mass accretion rate from the envelope can be written as $\dot{M}=c_{\rm eff}^3/G$,
where $G$ is the gravitational constant.

\blue{
At the vicinity of an accreting protostar, an accretion disk forms due to angular momentum of the inflowing material from large scales
\citep{2010MNRAS.403...45S,2011Sci...331.1040C,2011ApJ...737...75G}.}
The episodic nature of disk accretion decreases the time-averaged accretion rate onto the protostar as $\Mdot_{\star}=\eta \Mdot$, 
where we adopt $\eta=0.3$ as the conversion efficiency \blue{owing to angular momentum of the accreting flow} \citep{2016MNRAS.459.1137S, 2023MNRAS.518.1601T}. 
%\blue{
%  (Toyouchi+ find 0.4 actually.)
%}
The size evolution of an accreting protostar and the radiative feedback strength depend sensitively on whether
the accretion rate becomes higher than a critical value of $\dot{M}_{\mathrm{crit}}=0.04~\Msunyr$ 
\citep{2001ApJ...561L..55O,2013ApJ...778..178H,2013A&A...558A..59S,2015MNRAS.452..755S,2018MNRAS.474.2757H}.
For $\dot{M}_\star > \Mdot_\mathrm{crit}$, the stellar envelope inflates owing to rapid entropy injection through accreting matter,
keeping its surface temperature as low as $T_\mathrm{eff} \simeq 5000$ K. 
Since stellar UV radiation is hardly emitted from the cold surface, the accretion flow efficiently feeds the central star without being 
impeded by radiative feedback.
As a result, the stellar mass reaches $\sim 10^5~\Msun$ but is limited by the general-relativistic instability to
%
\begin{equation}
 M_{\star, \mathrm{GR}} \simeq\left[0.83~\log \left(\frac{\dot{M}_{\star}}{\Msunyr}\right)+2.48\right] \times 10^{5} ~\Msun,
\end{equation}
%
in the range of $\dot{M}_\star\gtrsim 0.1~\Msunyr$ \citep{2016PhRvD..94b1501S,2017ApJ...842L...6W,2019PASA...36...27W}.
On the other hand, for $\dot{M}_\star \leq \Mdot_\mathrm{crit}$, the protostar begins to contract by loosing its thermal energy via radiative diffusion
and increases its surface temperature to $T_\mathrm{eff} \sim 10^5$ K.
Hence, intense stellar UV radiation heats the disk surface and launches mass outflows, preventing mass supply from the disk to the star
\citep{2008ApJ...681..771M,2011Sci...334.1250H}. 
In this case, the stellar mass is determined by the balance between mass accretion and mass loss via photoevaporation \citep{2013ApJ...773..155T}.
The feedback-regulated mass can be approximated as
\footnote[2]{A recent study by \citet{2023MNRAS.518.1601T} provides a more sophisticated prescription of the feedback-regulated stellar mass based on their radiation hydrodynamical simulations.}
%
\begin{equation}
M_{\star, \mathrm{fb}} \simeq 2.9 \times 10^{3} \Msun\left(\frac{\dot{M}_{\star}}{0.01 M_{\odot} \mathrm{yr}^{-1}}\right),
\end{equation}
%
see more details in \cite{2021ApJ...917...60L}.
For an intermediate accretion rate between $\Mdot_\mathrm{crit}$ and $0.1~\Msunyr$, 
we perform logarithmic interpolation in the $\Mdot_\star - M_\star$ plane to smoothly connect 
$M_{\star, \mathrm{fb}}$ and $M_{\star, \mathrm{GR}}$ at the boundaries.
At the end of the stellar lifetime, those massive primordial stars result in BHs without significant mass loss 
via stellar winds \citep{2003ApJ...591..288H,2015MNRAS.451.4086S}.
\blue{One important caveat is the effect of stellar rotation on the structure evolution.
Namely, the mass loss rate of a rotating massive star via winds would be enhanced by the centrifugal force on the surface
or suppressed by efficient mixing of the interior structure by meridional circulation \citep[see][references therein]{2008A&A...489..685E,2012A&A...542A.113Y}.
Although those effects in the high mass regime are poorly understood yet, full general-relativistic simulations of the gravitational collapse of a rotating supermassive star 
show that most of the stellar mass is eventually swallowed by the newly born BH, ejecting only $\sim 10\%$ of the mass \citep{2002ApJ...572L..39S,2016PhRvD..94b1501S}.}
Therefore, we assume that the mass of a remnant BH is equal to that of its stellar progenitor.

@@

\vspace{2mm}
\subsection{BH mass growth}
\label{sec:model}
With BH seeds planted, we study the evolution of the BH mass distribution until $z=6$. 
The growth model of each BH is characterized by a minimal number of free parameters, 
giving the accretion rate by
%
\begin{equation}
  \label{eq:mdot}
  \Mdot_\bullet = \lambda f(\Mbh) \Mdot_\mathrm{Edd} ,
\end{equation}
where $\lambda$ is the ratio of the quasar bolometric luminosity to its Eddington luminosity $L_{\rm Edd}$,
% $\lambda \equiv L/L_\mathrm{Edd}$ 
and $\Mdot_\mathrm{Edd} \equiv L_{\mathrm{Edd}}/\eta_0 c^2$ is the Eddington accretion rate with a radiative efficiency of $\eta_0=0.1$ \citep{1973A&A....24..337S},
which is consistent with the value obtained from the Soltan-Paczy{\'n}ski argument \citep[e.g.,][]{2002MNRAS.335..965Y,2010ApJ...725..388C}.
Based on studies on the mass dependence of the radiative efficiency for AGNs at $z\lesssim$ 3 
\citep{2008MNRAS.390..561C,2012ApJ...749..187L,2014ApJ...786..104U}, 
we introduce a similar functional form of
%
\begin{equation}
\label{eq:f_M}
f(\Mbh) = \frac{2}{1+\left(\Mbh /M_{\bullet,c} \right)^\delta}, 
\end{equation}
%
where $M_{\bullet,c}=10^8~\Msun$ is adopted
\footnote[3]{
A series of radiation hydrodynamical simulations for BH accretion
suggest that the gas supplying rate from galactic scales into the nuclear scale is not high enough to sustain 
super-Eddington accretion when the BH mass is higher than the characteristic value \citep{2021ApJ...907...74T}.}.
With a positive value of $\delta$, the BH growth at the high mass end is suppressed,
while the growth speed for less massive BHs is accelerated.
In the limit of $\delta \ll 1$, where $f(\Mbh) \simeq 1$ is nearly independent of $M_{\bullet,c}$,
the model in Eq.~(\ref{eq:f_M}) reproduces an exponential growth with an $e$-folding timescale of
$t_{\rm S} =  \Mbh/\Mdot_{\mathrm{Edd}} \approx 45$ Myr (the so-called Salpeter timescale; \citealt{1964ApJ...140..796S}).
\blue{
Note that we do not explicitly consider the energy loss with radiation in Eq.~(\ref{eq:mdot}), neglecting a factor of $1-\eta_0=0.9$.
This small deviation from unity can be absorbed by the uncertainty due to the functional form of $f(M_\bullet)$ and does not bring 
a significant impact on the results discussed below.
}

@@

% obs fraction 
A widely-accepted mechanism causing quasar obscuration is extinction by dusty tori and dense gas clouds 
in the circumnuclear region, hence tightly related to gas fueling and feedback processes from accreting BHs 
\citep[see][for a review]{2018ARA&A..56..625H}.
In addition, dense gas clouds at larger galactic scales may have a significant contribution to quasar obscuration at 
higher redshifts \citep{2020MNRAS.495.2135N}.
%and the obscured fraction is expected to increase with redshift \citep[]{2022A&A...666A..17G}. 
%However, we adopt the prescription of the obscured fraction constrained from quasar observations at $z\lesssim 5$ 
However, the physical origin and conditions for high-$z$ quasar obscuration have been poorly understood yet.
Observationally, the obscured fraction of AGNs with hydrogen column density of $N_{\rm H}>10^{22}~{\rm cm}^{-2}$
can be estimated from spectral analyses in the X-ray band
\citep[e.g.,][]{2003ApJ...598..886U,2007A&A...463...79G,2008A&A...490..905H}. 
The obscured fraction is nearly $\gtrsim 80\%$ at $L_{\rm X}<10^{43}~{\rm erg~s}^{-1}$ and 
decreases with the X-ray luminosity \citep{2014ApJ...786..104U,2014MNRAS.437.3550M}.
The exact function shape still remains under debate, but the overall dependence on the X-ray luminosity holds.
Moreover, the obscured fraction is found to increase gradually toward higher redshifts and is saturated at $z\gtrsim 2-3$
\citep{2008A&A...490..905H,2014ApJ...786..104U,2014MNRAS.437.3550M,2018MNRAS.473.2378V,2022A&A...666A..17G}.

@@

From another point of view, the discrepancy between the UV and X-ray QLFs at the faint end would be explained by the fact that 
only bright quasars outshining their host galaxies can be identified as point-like sources in the rest-UV quasar selection 
(\citetalias{2018ApJ...869..150M}; \citealt{2020MNRAS.495.2135N,2020MNRAS.494.1771A,2021MNRAS.502.2757O,2021MNRAS.502..662B,2022AJ....164..114K}).
However, at the faint end of $\Muv\gtrsim -24$ mag, extended sources such as the brightest galaxies at $z\sim 6$ 
dominate over quasars in number at the same UV magnitude \citep{2022ApJS..259...20H}.
Thus, the current observations might miss quasars embedded in extended galaxy populations. 
We leave this issue as an important caveat in high-$z$ quasar observations.
